### 2021-10-13 chapter2/chapter.tex
easy parts of the first round of revisions on Ch 1/2 from Nigel

%!TEX root = ../thesis.tex
\chapter{A stochastic modelling approach to approximating fluid queues\label{sec: construction and modelling}}
In this chapter we develop a new discretisation of a fluid queue using a \emph{quasi-birth-and-death process with rational arrival process components} (QBD-RAP). In the following chapter, Chapter \ref{sec: conv} we prove the scheme converges weakly. To improve upon the discretisation of \cite{bo2013}, we argue that the shift from QBD to QBD-RAP is necessary. The approximation can be made arbitrarily accurate. Due to its stochastic interpretation, the QBD-RAP discretisation ensures solutions are positive, even when discontinuities are present. Unlike slope limiters and post-hoc filtering, for the QBD-RAP the positivity preserving nature is a property of the discretised operator.

@@

For a given \(\Delta\), the discretisation of Bean and O'Reilly supposes that, when the phase is \(i\) and on the event of no change of phase, the sojourn time in a given level has an exponential distribution with rate \(|c_i|\Delta\). However, the corresponding sojourn time of the fluid queue is deterministic: given \(X(t) = x \in\calD_k\), then \(\{X(t)\}\) will leave \(\calD_k\) in exactly \(((k+1)\Delta - x)/|c_i|\) units of time if \(c_i>0\), and \((x-k\Delta)/|c_i|\) units of time if \(c_i<0\), on the event that the phase does not change before this time. We can extend the QBD model of Bean and O'Reilly to model this determinism more accurately by supposing that the sojourn times in each interval have Erlang distributions (rather than exponential). However, we effectively realise the same QBD approximation as the original, but on a finer discretisation. For example, using Erlang distributions of order 2 we would get a generator of the QBD approximation
\[\bs B = \left[ \begin{array}{cccc} \ddots & \ddots & \ddots & \\ \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) & \\  & \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) \\ & \ddots & \ddots & \ddots \end{array} \right]\]
\begin{align}
\bs B_0 &= \bs T\otimes I_2 - \left[ \begin{array}{ccc} \left[\begin{array}{cc} -\Delta/2 & \Delta/2 \\ 0 & -\Delta/2 \end{array}\right] \otimes \bs C_+ & & \\ & \left[\begin{array}{cc} -\Delta/2 & 0 \\ \Delta/2 & -\Delta/2 \end{array}\right] \otimes C_- & \\ & & \bs 0 \end{array}\right], 
\end{align}
\begin{align}
\bs B_{-1} &= \left[\begin{array}{ccc} \bs 0 & & \\ & \bs C_-\otimes \left[\begin{array}{cc} 0 & \Delta/2 \\ 0 & 0 \end{array}\right] & \\ & & \bs 0 \end{array}\right],
&&\bs B_{+1} = \left[\begin{array}{ccc} \bs C_+\otimes \left[\begin{array}{cc} 0 & 0 \\ \Delta/2 & 0 \end{array}\right] & & \\ & \bs 0 & \\ & & \bs 0\end{array}\right].
\end{align}
But this is the same discretisation we would get if we were to use intervals of width \(\Delta/2\) in the QBD approximation of \cite{bo2013}, with the rows/columns reordered. Indeed, this appears to be the best we can do with a QBD-MAP approximation as the Erlang distribution is known to be the least variable Phase-type distribution for a given order \citep{as1987} and therefore gives the closest approximation to the deterministic behaviour we are trying to approximate. This necessitates the extension to QBD-RAPs, whereby we can find more concentrated distributions than any Phase-type, but retain (some of) the stochastic interpretation and matrix-analytic tools.

@@

\section{Time to exit an interval}\label{sec: modelling}
Consider partition of the state space of the level of a fluid queue, \([0,M]\) into \(K\) intervals of width \(\Delta = M/(K+1)\), specifically \(\mathcal D_{k,i} = [k\Delta,(k+1)\Delta)\) if \(i\in\mathcal S_+\) and \(\mathcal D_{k,i} = (k\Delta,(k+1)\Delta]\) if \(i\in\mathcal S_-\), \(k=0,\dots,K+1\). The distinction between \(\mathcal D_{k,i}\) for \(i\in\mathcal S_+\) and \(\calS_-\) is a technical one and will be discussed later. For now, one reason we might want to specify the discretisation in this way, is that it ensures the stochastic process 
\begin{align}
	\displaystyle \left\{\sum_{k\in\mathcal K}\sum_{i\in\calS} k1(X(t)\in\calD_{k,i})\right\}_{t\geq 0} \label{eqn: kfj}
\end{align}
is c\'adl\'ag. The expression (\ref{eqn: kfj}) is the discrete process of the fluid-queue which the level process of the QBD-RAP will approximate. Let \(y_k = k\Delta\), \(k=0,\dots,K+1\) and \(\calD_k = [y_k,y_{k+1}]\). We now look to model the behaviour of \(\{X(t)\}\) on an interval \(\calD_k\) over the time \([t,t+u]\) given information about the phase process over this time.

@@

To approximate the deterministic behaviour observed above, consider a matrix-exponential distribution \(Z\sim ME(\bs \alpha, \bs S, \bs s)\) with mean \(\Delta\) and low variance. As we shall see later, the assumption that \(Z\) has low variance will allow us to claim that \(Z\) can be used to approximate deterministic behaviour. Let \(Z_i\sim ME(\bs \alpha, \bs S_i, \bs s_i)\) where \(\bs S_i = |c_i|\bs S\), \(\bs s_i=-\bs S_i\bs e\), \(i\in\calS\). The random variable \(Z_i\) has mean \(\Delta/|c_i|\). Further, let \(R_i(u)\) be the residual time, \(R_i(u) = Z_i-u\mid \{Z_i-u>0\}\), \(i\in\calS\). Suppose first that \(X(t)=y_k\) and \(\varphi(t)=i\in\mathcal S_+\), so that \(X(t)\) will leave \(\mathcal D_{k,i}\) in exactly \(\Delta/|c_i|\) units of time on the event that the phase process remains in \(i\) for at least this amount of time.

After \(u\in[0,\Delta/|c_i|)\) amount of time has elapsed, on the event that \(\varphi(t+s)=i\), for all \(s\in[0,u)\), then \(X(t+u) = y_k + c_i u\). If the phase remains \(i\) for a further \(\Delta/|c_i|-u\) amount of time, then \(\{X(t)\}\) will leave \(\calD_{k,i}\) at exactly this time. Also at time \(t+u\), given \(Z_i>u\), the distribution of the residual time, \(R_i(u)\), has density 
\[f_{R_i(u)}(r) = \cfrac{\bs\alpha e^{\bs S_i(u+ r) } \bs s_i}{\bs\alpha e^{\bs S_iu } \bs e} = \bs A(t+u) e^{\bs S_i r}  \bs s_i.\]
Here, the event \(Z_i>u\) approximates the event \(X(t)\in\calD_{k,i}\) and we want the residual time, \(R_i(u)\), to approximate the time until \(\{X(t)\}\) leaves \(\calD_{k,i}\). That is, we want \(R_i(u)\) to approximate a deterministic random variable at \(\Delta/|c_i| - u\). To this end, we observe that, if the variance of \(Z_i\) is sufficiently low, then for \(\varepsilon>0\) and \(u<\Delta/|c_i|-\varepsilon\), then
\begin{align*}
	&\mathbb P(R_i(u) \in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon)) 
	\\&= \mathbb P(Z_i\in (\Delta/|c_i|-\varepsilon, \Delta/|c_i| + \varepsilon)\mid Z_i > u) 
	\\&= \cfrac{\mathbb P(Z_i\in (\Delta/|c_i|-\varepsilon, \Delta/|c_i| + \varepsilon))}{\mathbb P(Z_i > u)} 
	\\&= \cfrac{\mathbb P(Z\in (\Delta-\varepsilon, \Delta + \varepsilon))}{\mathbb P(Z_i > u)} 
%	\\&= \cfrac{\mathbb P(Z_i-u\in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))}{\mathbb P(Z_i-u > 0)} 
%	\\&\geq \mathbb P(Z-u\in (\Delta-u-\varepsilon, \Delta-u + \varepsilon))
	\\&\geq \cfrac{1-\cfrac{\var(Z)}{\varepsilon^2}}{1} 
\end{align*}
since, by Chebyshev's inequality, \(\mathbb P(Z_i\in (\Delta/|c_i|-\varepsilon, \Delta/|c_i| + \varepsilon)) \geq 1- \cfrac{\var(Z_i)}{\varepsilon^2}\). Choosing \(\varepsilon = \var(X)^{1/3}\), then 
\[\mathbb P(R_i(u) \in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))\geq 1-\var(Z)^{1/3} \approx 1. \]
That is, when the variance of \(Z_i\) is low, the residual time \(R_i(u)\) will be concentrated around \(\Delta/|c_i| - u\), as required. Figure \ref{fig:residual distributions} gives an example of a density function for a \(Z_i\) with mean \(\Delta=1\) and \(c_i=1\), as well as the density function of \(R_i(0.3)\) and \(R_i(0.5)\) for comparison. Observe that the density of the residual life \(R_i(0.3)\) is concentrated around \(\Delta-0.3 = 0.7\) and, similarly the density of the residual life \(R_i(0.6)\) is concentrated around \(\Delta-0.6 = 0.4\).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{chapter2/figs/ME_residual_life_density.pdf}
\caption{The density function for a \emph{concentrated matrix exponential of order 21} from \citep{hhat2020} (blue) and corresponding density functions of the residual lives, \(R_{0.3}\) (red), \(R_{0.5}\) (blue). Observe how the density function of the \(Z_i\) (blue) to approximate a point mass at \(\Delta=1\), while the density functions of \(R_{0.3}\) (red) and \(R_{0.5}\) (blue) approximate point masses at \(0.7\) and \(0.4\), respectively. }
\label{fig:residual distributions}
\end{figure}

@@

\paragraph{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\)} 
%Let \(Z_j\sim ME(\bs\alpha,\bs S_j)\) be a matrix exponential random variable with mean \(\Delta/|c_j|\) and let \(R_j(u)\) be the residual time \(R_j(u) = Z_j-u \mid\{Z_j-u>0\}\). %At time \(t+u\) the value of the orbit is \(\bs A(t+u) =  \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\), which can be equivalently written as \(\bs A(t+u) =  \cfrac{\bs\alpha e^{\bs S_ju/|c_i| } }{\bs\alpha e^{\bs S_ju/|c_i| } \bs e}\).
Consider the residual time \(R_j(|c_i|u/|c_j|)\), which is the distribution of time until \(Z_j\) occurs, given \(Z_j>|c_i|u/|c_j|\). The initial vector associated with the residual time \(R_j(|c_i|u/|c_j|)\) is \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\). This is the initial vector which is obtained if at time \(t\) the phase is \(\varphi(t)=j\) and the orbit is \(\bs A(t)=\alpha,\) so that the QBD-RAP has just entered level \(k\), and there is no change of phase by time \(|c_i|u/|c_j|\). Above, we argued that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) can be seen to correspond to the event that \(X(t+u)\) is \(|c_j|\left(|c_i|u/|c_j|\right) = |c_i|u\) units greater than \(y_k\). 
%We argued earlier that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) is the position of the orbit on the event that at time \(t\) there is a change of level so that \(\bs A(t)=\bs \alpha\), and the phase is \(\varphi(t)=j\), followed by no change of phase by time \(t+|c_i|u/|c_j|\). The analogous event of the fluid queue is that at time \(t\) the phase is \(\varphi(t)=j\), the fluid level is \(X(t)=y_k\), so that the fluid level has just entered an interval \(\calD_k\) at time \(t\), and there is no change of phase by time \(t+|c_i|u/|c_j|\), in which case the fluid level will be \(X(t+|c_i|u/|c_j|)=y_\ell + |c_j|\left(|c_i|u/|c_j|\right)=y_\ell + |c_i|u\). Therefore, we claim that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) is the correct position of the orbit to approximate the fact that \(\{X(t)\}\) is \(|c_i|u\) units to the right of the point \(y_k\).

@@

Further to this argument, observe that the residual time \(R_j(|c_i|u/|c_j|)\) has density 
\[f_{R_j(|c_i|u/|c_j|)}(r) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}e^{\bs S_j r}\bs s_j,\]
%and corresponds to an orbit position of \(\bs A(t+u) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\). 
and 
\begin{align}
	&\mathbb P(R_j(|c_i|u/|c_j|) \in ((\Delta-|c_i|u)/|c_j|-\varepsilon, (\Delta-|c_i|u)/|c_j| + \varepsilon)) 
	\\&=\mathbb P(Z_j\in (\Delta/|c_j|-\varepsilon, \Delta/|c_j| + \varepsilon)\mid Z_j > |c_i|u/|c_j|) 
	\\&= \cfrac{\mathbb P(Z_j\in (\Delta/|c_j|-\varepsilon, \Delta/|c_j| + \varepsilon))}{\mathbb P(Z_j > |c_i|u/|c_j|)} 
%	\\&= \cfrac{\mathbb P(Z_i-u\in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))}{\mathbb P(Z_i-u > 0)} 
%	\\&\geq \mathbb P(Z-u\in (\Delta-u-\varepsilon, \Delta-u + \varepsilon))
%	\\&\geq \cfrac{1-\cfrac{\var(Z_j)}{\varepsilon^2}}{1} 
	\\&\geq 1-\var(Z)^{1/3} \approx 1.
\end{align}
Hence, when the variance of \(Z\) is low, the residual time, \(R_j(|c_i|u/|c_j|)\), is concentrated around \((\Delta - |c_i|u)/|c_j|\) as required.

@@

\paragraph{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\)} Now suppose \(X(t)=y_k\), \(\varphi(t)=i\in\calS_+\), and the phase remains in state \(i\) until there is a change of phase phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), \(u\in[0,\Delta/|c_i|)\). We have reasoned, intuitively, that, for \(i\in\calS_+\) and for \(x=|c_i|u\in[0,\Delta)\), the orbit position \(\cfrac{\bs\alpha e^{\bs Sx} }{\bs\alpha e^{\bs Sx } \bs e}\) corresponds to the fluid level position \(X(t) = y_k+x\), that is \(x\) units greater than \(y_k\). Meanwhile, the residual time \(R_i(u)\) \emph{approximates} the fact that \(X(t)\) is \(\Delta-x\) units to the left of \(y_{k+1}\). We have also reasoned that for \(i\in\calS_-\) the orbit position \(\cfrac{\bs\alpha e^{\bs Sx} }{\bs\alpha e^{\bs Sx } \bs e}\) corresponds to the fluid level position \(X(t) = y_{k+1}-x\), \(x\) units less than \(y_{k+1}\), and the residual time \(R_i(u)\) \emph{approximates} the fact that \(X(t)\) is \(\Delta-x\) units greater than \(y_k\). Notice that in positive phases the orbit position describes the exact position of \(X(t)\) from the left of the interval but only approximates the position of \(X(t)\) from the right of the interval, and vice-versa for negative phases.

To construct the QBD-RAP we need to find a matrix \(\bs D\), say, to take the information about the position of \(X(t)\) contained in the orbit when in positive phases, and map it to information about the position of \(X(t)\) in negative phases. Then, when there is a change of phase at time \(t+u\) from \(i\in\calS_+\) to \(j\in\calS_-\), the orbit position will jump from \(\bs A(t+u^-)\) to \(\bs A(t+u)=\bs A(t+u^-)\bs D\). The matrix \(\bs D\) must have the property that, for any \(\bs a\in\mathcal A\), \((\bs a\bs D, \bs S)\) is a representation of a matrix exponential distribution. The matrix \(\bs D\) should also have the property that \(\bs D \bs e = \bs e\) as this will mean that we can model the phase process of the fluid exactly.

>>>>

%!TEX root = ../thesis.tex
\chapter{A stochastic modelling approach to approximating fluid queues\label{sec: construction and modelling}}
In this chapter we develop a new discretisation of a fluid queue using a \emph{quasi-birth-and-death process with rational arrival process components} (QBD-RAP). In the following chapter, Chapter \ref{sec: conv} we prove the scheme converges weakly. To improve upon the discretisation of \cite{bo2013}, we argue that the shift from QBD to QBD-RAP is necessary. The QBD construction of \cite{bo2013} uses Erlang distributions to approximate the deterministic time that the fluid queue spends in an interval on the event that the phase is constant -- it is well known that the Erlang distribution is the distribution with the smallest variance in the class of Phase-type distributions of a given order \cite{as1987}, and in this sense is the best approximation to the determinstic behaviour we want to capture.

By choosing a matrix-exponential distribution with sufficiently small variance, the QBD-RAP approximation can be made arbitrarily accurate. Due to its stochastic interpretation, the QBD-RAP discretisation ensures solutions are positive, even when discontinuities are present. Unlike slope limiters and post-hoc filtering, for the QBD-RAP the positivity preserving nature is a property of the discretised operator.

@@

For a given \(\Delta\), the discretisation of Bean and O'Reilly supposes that, when the phase is \(i\) and on the event of no change of phase, the sojourn time in a given level has an exponential distribution with rate \(|c_i|\Delta\). However, the corresponding sojourn time of the fluid queue is deterministic: given \(X(t) = x \in\calD_k\), then \(\{X(t)\}\) will leave \(\calD_k\) in exactly \(((k+1)\Delta - x)/|c_i|\) units of time if \(c_i>0\), and \((x-k\Delta)/|c_i|\) units of time if \(c_i<0\), on the event that the phase does not change before this time. We can extend the QBD model of Bean and O'Reilly to model this determinism more accurately by supposing that the sojourn times in each interval have Erlang distributions (rather than exponential). However, we effectively realise the same QBD approximation as the original, but on a finer discretisation. For example, using Erlang distributions of order 2 we would get a generator of the QBD approximation
\[\bs B = \left[ \begin{array}{cccc} \ddots & \ddots & \ddots & \\ \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) & \\  & \bs B_{-1}(\Delta) & \bs B_0(\Delta) & \bs B_{+1}(\Delta) \\ & \ddots & \ddots & \ddots \end{array} \right]\]
\begin{align}
\bs B_0 &= \bs T\otimes I_2 - \left[ \begin{array}{ccc} \left[\begin{array}{cc} -\Delta/2 & \Delta/2 \\ 0 & -\Delta/2 \end{array}\right] \otimes \bs C_+ & & \\ & \left[\begin{array}{cc} -\Delta/2 & 0 \\ \Delta/2 & -\Delta/2 \end{array}\right] \otimes C_- & \\ & & \bs 0 \end{array}\right], 
\end{align}
\begin{align}
\bs B_{-1} &= \left[\begin{array}{ccc} \bs 0 & & \\ & \bs C_-\otimes \left[\begin{array}{cc} 0 & \Delta/2 \\ 0 & 0 \end{array}\right] & \\ & & \bs 0 \end{array}\right],
&&\bs B_{+1} = \left[\begin{array}{ccc} \bs C_+\otimes \left[\begin{array}{cc} 0 & 0 \\ \Delta/2 & 0 \end{array}\right] & & \\ & \bs 0 & \\ & & \bs 0\end{array}\right].
\end{align}
But this is the same discretisation we would get if we were to use intervals of width \(\Delta/2\) in the QBD approximation of \cite{bo2013}, with the rows/columns reordered. Indeed, this appears to be the best we can do with a QBD approximation as the Erlang distribution is known to be the least variable Phase-type distribution for a given order \citep{as1987} and therefore gives the closest approximation to the deterministic behaviour we are trying to approximate. This necessitates the extension to QBD-RAPs, whereby we can find more concentrated distributions than any Phase-type and retain enough stochastic interpretation and matrix-analytic tools.

@@

\section{Time to exit an interval}\label{sec: modelling}
Consider the partition of the state space of the level of a fluid queue, \([0,M]\) into \(K\) intervals of width \(\Delta = M/(K+1)\), specifically \(\mathcal D_{k,i} = [k\Delta,(k+1)\Delta)\) if \(i\in\mathcal S_+\) and \(\mathcal D_{k,i} = (k\Delta,(k+1)\Delta]\) if \(i\in\mathcal S_-\), \(k=0,\dots,K+1\). The distinction between \(\mathcal D_{k,i}\) for \(i\in\mathcal S_+\) and \(\calS_-\) is a technical one and will be discussed later. For now, one reason we might want to specify the discretisation in this way, is that it ensures the stochastic process 
\begin{align}
	\displaystyle \left\{\sum_{k\in\mathcal K}\sum_{i\in\calS} k1(X(t)\in\calD_{k,i})\right\}_{t\geq 0} \label{eqn: kfj}
\end{align}
is right-continuous with left limits (c\'adl\'ag). The expression (\ref{eqn: kfj}) is the discrete process of the fluid-queue which the level process of the QBD-RAP will approximate. Let \(y_k = k\Delta\), \(k=0,\dots,K+1\) and \(\calD_k = [y_k,y_{k+1}]\). We now look to model the behaviour of \(\{X(t)\}\) on an interval \(\calD_k\) over the time \([t,t+u]\) given information about the phase process over this time.

@@

To approximate the deterministic behaviour observed above, consider a matrix-exponential distribution \(Z\sim ME(\bs \alpha, \bs S, \bs s)\) with mean \(\Delta\) and low variance. As we shall see later, the assumption that \(Z\) has low variance will allow us to claim that \(Z\) can be used to approximate deterministic behaviour. Let \(Z_i\sim ME(\bs \alpha, \bs S_i, \bs s_i)\) where \(\bs S_i = |c_i|\bs S\), \(\bs s_i=-\bs S_i\bs e\), \(i\in\calS\). The random variable \(Z_i\) has mean \(\Delta/|c_i|\). Further, let \(R_i(u)\) be the residual time, \(R_i(u) = (Z_i-u)1\{Z_i-u>0\}\), \(i\in\calS\). Suppose first that \(X(t)=y_k\) and \(\varphi(t)=i\in\mathcal S_+\), so that \(X(t)\) will leave \(\mathcal D_{k,i}\) in exactly \(\Delta/|c_i|\) units of time on the event that the phase process remains in \(i\) for at least this amount of time.

After \(u\in[0,\Delta/|c_i|)\) amount of time has elapsed, on the event that \(\varphi(t+s)=i\), for all \(s\in[0,u)\), then \(X(t+u) = y_k + c_i u\). If the phase remains \(i\) for a further \(\Delta/|c_i|-u\) amount of time, then \(\{X(t)\}\) will leave \(\calD_{k,i}\) at exactly this time. Also at time \(t+u\), given \(Z_i>u\), the distribution of the residual time, \(R_i(u)\), has density 
\[f_{R_i(u)}(r) = \cfrac{\bs\alpha e^{\bs S_i(u+ r) } \bs s_i}{\bs\alpha e^{\bs S_iu } \bs e} = \bs A(t+u) e^{\bs S_i r}  \bs s_i.\]
Here, the event \(Z_i>u\) approximates the event \(X(t+u)\in\calD_{k,i}\) and we want the residual time, \(R_i(u)\), to approximate the time until \(\{X(t)\}\) leaves \(\calD_{k,i}\). That is, we want \(R_i(u)\) to approximate a deterministic random variable at \(\Delta/|c_i| - u\). To this end, we observe that for \(\varepsilon>0\) and \(u<(\Delta-\varepsilon)/|c_i|\), then
\begin{align*}
	&\mathbb P(R_i(u) \in ((\Delta-\varepsilon)/|c_i|-u, (\Delta+\varepsilon)/|c_i|-u) ) 
	\\&= \mathbb P(Z_i\in ((\Delta-\varepsilon)/|c_i|, (\Delta+\varepsilon)/|c_i|) ) 
	\\&= \mathbb P(Z\in (\Delta-\varepsilon, \Delta + \varepsilon))
%	\\&= \cfrac{\mathbb P(Z_i-u\in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))}{\mathbb P(Z_i-u > 0)} 
%	\\&\geq \mathbb P(Z-u\in (\Delta-u-\varepsilon, \Delta-u + \varepsilon))
	\\&\geq 1-\cfrac{\var(Z)}{\varepsilon^2}
\end{align*}
since, by Chebyshev's inequality, \(\mathbb P(Z\in (\Delta-\varepsilon, \Delta + \varepsilon)) \geq 1- \cfrac{\var(Z)}{\varepsilon^2}\). Choosing \(\varepsilon = \var(Z)^{1/3}\), then 
\[\mathbb P(R_i(u) \in ((\Delta-\varepsilon)/|c_i|-u, (\Delta+\varepsilon)/|c_i|) )\geq 1-\var(Z)^{1/3} \approx 1, \]
when \(\var(Z)\) is small. 
That is, when the variance of \(Z\) (equivalently \(Z_i\)) is low, the residual time \(R_i(u)\) will be concentrated around \(\Delta/|c_i| - u\), as required. Figure \ref{fig:residual distributions} gives an example of a density function for a \(Z_i\) with mean \(\Delta=1\) and \(c_i=1\), as well as the density function of \(R_i(0.3)\) conditional on \(Z_i>0.3\) and \(R_i(0.6)\) conditional on \(Z_i>0.6\), for comparison. Observe that the density of the residual life \(R_i(0.3)\) conditional on \(Z_i>0.3\) is concentrated around \(\Delta-0.3 = 0.7\) and, similarly the density of the residual life \(R_i(0.6)\) conditional in \(Z_i>0.6\) is concentrated around \(\Delta-0.6 = 0.4\).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{chapter2/figs/ME_residual_life_density.pdf}
\caption{The density function for a \emph{concentrated matrix exponential of order 21} from \citep{hhat2020} (blue) and corresponding density functions of the residual lives, \(R_{0.3}\) (red), \(R_{0.6}\) (blue). Observe how the density function of the \(Z_i\) (blue) to approximate a point mass at \(\Delta=1\), while the density functions of \(R_{0.3}\) (red) and \(R_{0.6}\) (blue) approximate point masses at \(0.7\) and \(0.4\), respectively. }
\label{fig:residual distributions}
\end{figure}

@@

\paragraph{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_+\)} 
%Let \(Z_j\sim ME(\bs\alpha,\bs S_j)\) be a matrix exponential random variable with mean \(\Delta/|c_j|\) and let \(R_j(u)\) be the residual time \(R_j(u) = Z_j-u \mid\{Z_j-u>0\}\). %At time \(t+u\) the value of the orbit is \(\bs A(t+u) =  \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\), which can be equivalently written as \(\bs A(t+u) =  \cfrac{\bs\alpha e^{\bs S_ju/|c_i| } }{\bs\alpha e^{\bs S_ju/|c_i| } \bs e}\).
Consider the residual time \(R_j(|c_i|u/|c_j|)\), which is the distribution of time until \(Z_j-|c_i|u/|c_j|\) occurs, on the event \(Z_j>|c_i|u/|c_j|\). The initial vector associated with the residual time \(R_j(|c_i|u/|c_j|)\) is \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\). This is the initial vector which is obtained if at time \(t\) the phase is \(\varphi(t)=j\) and the orbit is \(\bs A(t)=\bs \alpha,\) so that the QBD-RAP has just entered level \(k\), and there is no change of phase by time \(|c_i|u/|c_j|\). Above, we argued that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) can be seen to correspond to the event that \(X(t+u)\) is \(|c_j|\left(|c_i|u/|c_j|\right) = |c_i|u\) units greater than \(y_k\). 
%We argued earlier that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) is the position of the orbit on the event that at time \(t\) there is a change of level so that \(\bs A(t)=\bs \alpha\), and the phase is \(\varphi(t)=j\), followed by no change of phase by time \(t+|c_i|u/|c_j|\). The analogous event of the fluid queue is that at time \(t\) the phase is \(\varphi(t)=j\), the fluid level is \(X(t)=y_k\), so that the fluid level has just entered an interval \(\calD_k\) at time \(t\), and there is no change of phase by time \(t+|c_i|u/|c_j|\), in which case the fluid level will be \(X(t+|c_i|u/|c_j|)=y_\ell + |c_j|\left(|c_i|u/|c_j|\right)=y_\ell + |c_i|u\). Therefore, we claim that the vector \(\cfrac{\bs \alpha e^{\bs S_j |c_i|u/|c_j| } }{\bs \alpha e^{\bs S_j |c_i|u/|c_j|} \bs e }\) is the correct position of the orbit to approximate the fact that \(\{X(t)\}\) is \(|c_i|u\) units to the right of the point \(y_k\).

@@

Further to this argument, observe that the residual time \(R_j(|c_i|u/|c_j|)\) has density 
\[f_{R_j(|c_i|u/|c_j|)}(r) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}e^{\bs S_j r}\bs s_j,\]
%and corresponds to an orbit position of \(\bs A(t+u) = \cfrac{\bs\alpha e^{\bs S_iu } }{\bs\alpha e^{\bs S_iu } \bs e}\). 
and 
\begin{align}
	&\mathbb P(R_j(|c_i|u/|c_j|) \in ((\Delta-|c_i|u-\varepsilon)/|c_j|, (\Delta-|c_i|u + \varepsilon)/|c_j| )) \nonumber
	\\&=\mathbb P(Z_j\in ((\Delta-\varepsilon)/|c_j|, (\Delta + \varepsilon)/|c_j|)) 
	\\&=\mathbb P(Z\in (\Delta-\varepsilon ,  \Delta + \varepsilon )) 
	% \\&=  \mathbb P(Z_j\in (\Delta/|c_j|-\varepsilon, \Delta/|c_j| + \varepsilon)) 
%	\\&= \cfrac{\mathbb P(Z_i-u\in (\Delta/|c_i|-u-\varepsilon, \Delta/|c_i|-u + \varepsilon))}{\mathbb P(Z_i-u > 0)} 
%	\\&\geq \mathbb P(Z-u\in (\Delta-u-\varepsilon, \Delta-u + \varepsilon))
%	\\&\geq \cfrac{1-\cfrac{\var(Z_j)}{\varepsilon^2}}{1} 
	\\&\geq 1-\var(Z)^{1/3} \approx 1.
\end{align}
Hence, when the variance of \(Z\) is low, the residual time, \(R_j(|c_i|u/|c_j|)\), is concentrated around \((\Delta - |c_i|u)/|c_j|\) as required.

@@

\paragraph{On a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\)} Now suppose \(X(t)=y_k\), \(\varphi(t)=i\in\calS_+\), and the phase remains in state \(i\) until there is a change of phase from \(i\in\calS_+\) to \(j\in\calS_-\) at time \(t+u\), \(u\in[0,\Delta/|c_i|)\). We have reasoned, intuitively, that, for \(i\in\calS_+\) and for \(x=|c_i|u\in[0,\Delta)\), the orbit position \(\cfrac{\bs\alpha e^{\bs Sx} }{\bs\alpha e^{\bs Sx } \bs e}\) corresponds to the fluid level position \(X(t) = y_k+x\), that is \(x\) units greater than \(y_k\). Correspondingly, the residual time \(R_i(u)\) \emph{approximates} the fact that \(X(t)\) is \(\Delta-x\) units less than \(y_{k+1}\). We have also reasoned that for \(i\in\calS_-\) the orbit position \(\cfrac{\bs\alpha e^{\bs Sx} }{\bs\alpha e^{\bs Sx } \bs e}\) corresponds to the fluid level position \(X(t) = y_{k+1}-x\), that is \(x\) units less than \(y_{k+1}\), and the residual time \(R_i(u)\) \emph{approximates} the fact that \(X(t)\) is \(\Delta-x\) units greater than \(y_k\). Notice that in positive phases the orbit position describes the exact position of \(X(t)\) from the left-hand boundary of the interval but only approximates the position of \(X(t)\) from the right-hand boundary of the interval, and vice-versa for negative phases.

To construct the QBD-RAP we need to find a matrix \(\bs D\) such that when there is a change of phase at time \(t+u\) from \(i\in\calS_+\) to \(j\in\calS_-\), the orbit position will jump from \(\bs A(t+u^-)\) to \(\bs A(t+u)=\bs A(t+u^-)\bs D\). We want \(\bs A(t+u)\) to represent, in some way, a distance \(|c_i|u\) from \(y_k\) when viewed from a negative phase. The matrix \(\bs D\) must have the property that, for any \(\bs a\in\mathcal A\), \((\bs a\bs D, \bs S)\) is a representation of a matrix exponential distribution. We would also like the matrix \(\bs D\)to the property \(\bs D \bs e = \bs e\) as we can use this to show that phase process of the QBD-RAP and fluid queue have the same probability law (distribution).
